Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-12-13 paper/manuscript.tex
added todo
added best practices on data representation

\subsubsection*{Good analysis practices}
\todo[inline, color={green!20}]{ASJSM: I would quite like a summary of, here is how you should plot your free energy data, e.g.: \url{http://practicalcheminformatics.blogspot.com/2019/02/some-thoughts-on-evaluating-predictive.html}}

@@

%%%%%%%%%%%%%%%%%%%%
%              Terminology                  %
%%%%%%%%%%%%%%%%%%%%
\section{Terminology and abbreviations}
\label{sec:tem-abbrev}
\begin{itemize}
\item Feature Box covering major technical terms and abbreviations
\item Examples:
\begin{itemize}
\item EXP, BAR, MBAR
\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
\item FEP (free energy perturbation), alchemical, AFE (alchemical free energy)
\end{itemize}
\end{itemize}
%%%%%%%%%%%%%%%%%%%%
%                Software                    %
%%%%%%%%%%%%%%%%%%%%
\section{Available software -- a summary}
\todo[inline]{ASJSM: Expand and add more references. }
\label{sec:software}
There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
\begin{itemize}
\item Commercial:
   \begin{itemize}
    \item FEP+ is a tool offered by Schr\"{o}dinger Inc. under a commercial license. It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results. It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs. 
    \end{itemize}
\item Free or low-cost for academics / commercial for industry:
	\begin{itemize}
	\item CHARMM has a variety of tools developed over the years. PERT module can be used to define initial and final states and define the intermediate lambda points. FREN and BAR modules can be used to analyze the data after the MD run. Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.  
	\item TIES and AMBER FEW? (Peter Coveney)
	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
	\end{itemize}
\item Free (libre) open source:
	\begin{itemize}
	\item PLUMED is an open source tool which enables the usage of a variety of MD engines. It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly. It also offers a VMD based plugin for the computation of collective variables.   	
	\item SIRE  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design. 
	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
	\item GROMACS is a molecular simulation package with a significant number of free energy methods implementations. The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{lemkul2018From}.
	\item pmx for mutations
	\end{itemize}
\item Setup tools
	\begin{itemize}
	\item FESetup: AMBER, gromacs, Sire
	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM scripts for solvation and ligand-binding free energy calculations.
	\end{itemize}
\item Analysis tools:
	\begin{itemize}
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
	\url{https://github.com/michellab/freenrgworkflows}
	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
	\url{https://github.com/choderalab/pymbar}
	\end{itemize}
\end{itemize}

>>>>

\subsection*{Best practices for data representation}

Practitioners in the field are encouraged to follow best practices on analysis of results and data depictions. The following points should be considered when plotting alchemical free energy prediction results.
\todo[inline]{JS: @everyone I suspect people will have strong opinions on this subject, feel free to include your own suggestions}
\paragraph{Statistics to include}
As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in FE predictions. First and foremost, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph). It is recommended to perform triplicates of your predictions at minimum to ensure some measure of reliability in your data.

Because of the strong drug discovery context and the relevance of alchemical FE predictions to SAR projects, both the overall correlation of your predictions to experiment and the ability of your models to detect the best-scoring compounds (i.e. rankings) should be computed. Conventionally, this means including an R\textsuperscript{2} (or Pearson r) and a Kendall \texttau{} metric in your results. Additionally, practitioners may choose to include a Spearman \textrho{} as well.

Mean unsigned error (MUE, also called mean absolute error/MAE) is the final statistic to include in your results. Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still be multiple kcal/mol, debunking model performance. Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size. Even though this section may serve as an introduction to the main statistical practices, the authors recommend further reading on evaluation of computational models (cite Nicholls pt1 and pt2 + Jain&Nicholls "Recommendations for..").

\paragraph{Presenting your data}
Seeing as essentially all alchemical FE prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}). Most alchemical FE projects will concern 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars). Any study with \textgreater50 ligands should be trellised to not overestimate statistics. In that respect, it is bad practice to place multiple datasets on the same plot because this can suggest high model accuracy even though the individual models perform less well (cite blogpost Pat Walters).

Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e. $x=y$). Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions. These regions can serve as tools to communicate your model performance: any perturbation inside the 1 kcal/mol region can be seen as highly reliable, any perturbation inside the 2 kcal/mol region should be seen as somewhat reliable, and any perturbation outside the confidence regions should be expected to be unreliable (and handled as outliers). In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas. 
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig13_analysis_practices/Figure.pdf}
    \caption{Example of a well-depicted alchemical free energy prediction showing the relation between predicted and experimentally-determined Gibbs free energy in kcal per mol with standard errors as error bars. The two outliers (underestimations) are shown in red; the dark and light-orange regions depict the 1- and 2-kcal per mol confidence bounds.}
    \label{fig:scatterplot_analysis}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%
%              Terminology                  %
%%%%%%%%%%%%%%%%%%%%
\section{Terminology and abbreviations}
\label{sec:tem-abbrev}
\begin{itemize}
\item Feature Box covering major technical terms and abbreviations
\item Examples:
\begin{itemize}
\item EXP, BAR, MBAR
\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
\item FEP (free energy perturbation), alchemical, AFE (alchemical free energy)
\end{itemize}
\end{itemize}
\todo[inline]{JS: should this be combined with the acronyms section at the beginning of the paper?}
%%%%%%%%%%%%%%%%%%%%
%                Software                    %
%%%%%%%%%%%%%%%%%%%%
\section{Available software -- a summary}
\todo[inline]{ASJSM: Expand and add more references. }
\label{sec:software}
There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
\begin{itemize}
\item Commercial:
   \begin{itemize}
    \item FEP+ is a tool offered by Schr\"{o}dinger Inc. under a commercial license. It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results. It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs. 
    \end{itemize}
\item Free or low-cost for academics / commercial for industry:
	\begin{itemize}
	\item CHARMM has a variety of tools developed over the years. PERT module can be used to define initial and final states and define the intermediate lambda points. FREN and BAR modules can be used to analyze the data after the MD run. Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.  
	\item TIES and AMBER FEW? (Peter Coveney)
	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
	\end{itemize}
\item Free (libre) open source:
	\begin{itemize}
	\item PLUMED is an open source tool which enables the usage of a variety of MD engines. It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly. It also offers a VMD based plugin for the computation of collective variables.   	
	\item SIRE  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design. 
	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
	\item GROMACS is a molecular simulation package with a significant number of free energy methods implementations. The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{lemkul2018From}.
	\item pmx for mutations
	\end{itemize}
\item Setup tools
	\begin{itemize}
	\item FESetup: AMBER, gromacs, Sire
	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM scripts for solvation and ligand-binding free energy calculations.
	\end{itemize}
\item Analysis tools:
	\begin{itemize}
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
	\url{https://github.com/michellab/freenrgworkflows}
	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
	\url{https://github.com/choderalab/pymbar}
	\end{itemize}
\end{itemize}
